### 2018-07-17 report.tex
Backup

\documentclass[11pt, a4paper]{report}
% FOR COLORFUL TABLES
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{pdfpages}
\usepackage{array} % to make table column italic
\usepackage{textcomp} % to support the trademark symbol

@@

@@

@@

@@

\subsection{Decoy database preparation and search}

@@

\begin{table}[H]
\begin{tabular}{ccccc}
  \toprule
 Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
  \midrule
 46595 & 27384 & 19211 & 25056 & 10039 \\
 \bottomrule
\end{tabular}
\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
\label{tab:quantification_table}
\end{table}

A total of 46595 peptides were inferred to be present in at least one of the fractions. Of them more than 27 thousand mapped to a unique protein. However, a significant amount mapped to more than one protein. Such peptides are called non-unique peptides. The 25056 detected proteins were grouped into 10039 protein groups. Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them. Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from. Protein grouping decreased the amount of  considered protein entities to 10039.

@@

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L. During quantification, it is important to take into account as many of the effects influencing protein quantification. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

@@

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1.

@@

The low attained matching rate manifests existing room for improvement in the currently available tools. Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats. Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}. Their usage is predicted tor improve identification rates and quality of the matches.

@@

@@

\subsection{Bayesian statistics}

@@

The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data. While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data. Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way. This is the so called Bayes-theorem.

@@

The Bayesian framework enunciated above can be applied to the quantification problem. All is needed is the computation of the posterior probability distribution. Unfortunately, this is tractable analytically for simple models only. However, Markov Chain-Monte Carlo methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough. Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible. Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3), all making use of a programming paradigm called \textbf{probabilistic programming}.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{graph}
\end{figure}

>>>>

\documentclass[11pt, a4paper]{report}
% FOR COLORFUL TABLES
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{pdfpages}
\usepackage{array} % to make table column italic
\usepackage{textcomp} % to support the trademark symbol
\usepackage{minted}

@@

\DeclareAcronym{MCMC}{
  short = MCMC,
  long  = Markov Chain Monte Carlo,
  class = abbrev
}

@@

\DeclareAcronym{DAG}{
  short = DAG ,
  long  = Directed Acyclic Graph,
  class = abbrev
}

@@

\DeclareAcronym{ADVI}{
  short = ADVI ,
  long  = Automatic Differentiation Variational Inference,
  class = abbrev
}

\DeclareAcronym{NUTS}{
  short = NUTS ,
  long  = No-U-Turn-Sampler,
  class = abbrev
}

@@

\subsection{Decoy database preparation and search}
\label{subsec:database_preparation}

@@

\begin{table}[H]
\begin{tabular}{ccccc}
  \toprule
 Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
  \midrule
 46595 & 27384 & 19211 & 25056 & 10039 \\
 \bottomrule
\end{tabular}
\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
\label{tab:protein_inference}
\end{table}

A total of 46595 peptides were inferred to be present in at least one of the fractions (see table \ref{tab:protein_inference}). Of them more than 27 thousand mapped to a unique protein. However, a significant amount mapped to more than one protein. Such peptides are called non-unique peptides. The 25056 detected proteins were grouped into 10039 protein groups. Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them. Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from. Protein grouping decreased the amount of  considered protein entities to 10039.

@@

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L. During quantification, it is important to take into account as many of the effects influencing protein quantification. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

@@

In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A). Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins. The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too. Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power. Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected. Even though the observed correlation was low, the estimated log2FC was always greater than 1. The MQ+Rob pipeline systematically estimated log2FV values greater than the other 2.

@@

The low attained matching rate manifests existing room for improvement in the currently available tools. As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats. Only MS-GF+ was used in thi work for simplicity. Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}. Their usage is predicted to improve identification rates and quality of the matches.

@@

Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.

@@

\subsection{Frequentist and Bayesian statistics}

@@

The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data. While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data. Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way. This is the so called Bayes\textquotesingle s theorem.

@@

The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterir probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}

It can be applied to the quantification problem, where $\theta$ turns into the log2FC parameter we try to estimate. All is needed is the computation of the posterior probability distribution. Unfortunately, this is tractable analytically for simple models only. However, Markov Chain-Monte Carlo (\ac{MCMC}) methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough. Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible. Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3 \cite{Salvatier2016}), all making use of a programming paradigm called \textbf{probabilistic programming}.

@@

The management of uncertainty in statistics is done by means of probability distributions, which account for the possible values that a parameter could take, together with how likely they are. Probabilistic programming offers the framework to build complex statistical models by storing probability distributions as variables of the program. Moreover, probabilistic programming packages supply the tools needed to perform inference with these models by taking experimental evidence (data) and fitting the distributions to the data using Bayes theorem. The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using \ac{MCMC} algorithms provided with the probabilistic programming tool.

Pymc3 is a mature Python module dedicated to support probabilistic programming. It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (\ac{ADVI}) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}. Therefore, it provides the tools required to build a Bayesian model for the probabilisitc estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.

\item Develop a Bayesian model to estimate log2FC in proteomics datasets together with the uncertainty of the estimate.

\item Make it scalable and fast for usability in ordinary modern computers.

\subsection{Data input}
The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant} and \texttt{MSnSet2protdata} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.

\subsection{Sequence modelling}

When running with active sequence modelling, the peptide sequences and their neighborhood in the protein was extracted from \textit{E. coli} and \textit{Homo sapiens} proteome databases (\ref{subsec:database_preparation}). The protein neighborhood was defined by a window spanning 15 aminoacids on both sides of the peptide. It was extracted using the seqinr \cite{Charif2007} and GenomicRanges \cite{Lawrence2013} packages implemented in R and Bioconductor.

Features to model the peptide effect based on sequence were extracted using BLABLA

\subsection{Bayesian modelling using Pymc3}

The model was initialized

\begin{minted}[mathescape,
               linenos,
               xleftmargin=-70pt,
               numbersep=10pt,
               gobble=0,
               frame=lines,
               framesep=2mm]{python}
with pm.Model() as model:
\end{minted}

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{graph}
\caption{Diagram of the Bayesian model developed in the present paper, without sequence modelling of the peptide effect. It is represented as a Directed Acyclic Graph \ac{DAG} where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them. The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to. The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.}
\label{fig:daft_model}
\end{figure}

\subsection{Model implementation (I) peptide as a random process}

\subsection{Model implementation (II) peptide effect with sequence features}

\subsection{Posterior probability distribution computation}

\ac{ADVI} returns good results faster than \ac{NUTS}

\subsection{Model checking}
